Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-08-03 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Work on draft

# Pull in rain data and limit to storm-county pairs with 150 mm or more 
# of rainfall over lag -5 to 3. Figure out the cumulative values at each 
# day within a county-storm for each lag day.
rain_lag <- rbind(rain, lag_6) %>%
        arrange(storm_id, fips, lag) %>%
        mutate(storm_county = paste0(storm_id, "_", fips)) %>%
        filter(storm_county %in% storm_lag$storm_county) %>%
        group_by(storm_county) %>%
        mutate(cumulative_rain = cumsum(precip),
               next_lag = lead(lag),
               next_cumulative_rain = lead(cumulative_rain),
               over_50 = lead(precip) >= 50) %>%
        filter(lag != 3)
high_rain <- rain_lag %>%
        ungroup() %>%
        group_by(lag) %>%
        summarize(n_high = sum(over_50)) %>%
        ungroup() %>%
        mutate(lag = lag + 1)
to_highlight <- data.frame(x = c(-3.1, -3.1, 1.1, 1.1) + 0.5, 
                           y = c(-15, 500, 500, -15))

@@

```{r daily_rain, echo = FALSE, fig.width = 6, fig.height = 4}
rain_lag %>%
        filter(lag != -6) %>%
        ggplot(aes(x = lag, y = precip, group = lag)) + 
        geom_boxplot(outlier.size = 0.5) + theme_few() + 
        xlab("Lag period from storm's closest approach to county") + 
        ylab("Daily rainfall in county (mm)") 
```

@@

We compared exposure results based on the NLDAS-2 precipitation data with values from monitors reported for certain storms in the *Monthly Weather Reviews* yearly summary of the Atlantic basin tropical storm season. Here are some of the plots comparing county-level estimates of rain exposure from the NLDAS-2 data to all *MWR*-reported rainfall estimates within the county:

```{r echo = FALSE, fig.height = 4, fig.width = 6}
load("alex_gauge.Rdata")
alex_nldas <- county_rain(counties = unique(alex_gauge$fips),
                             start_year = 2010, end_year = 2010,
                             rain_limit = 0, dist_limit = 1500,
                             days_included = c(-2, -1, 0, 1))
alex_nldas <- alex_nldas %>%
        filter(storm_id == "Alex-2010")
alex_gauge$fips <- as.character(alex_gauge$fips)
compare_rain <- right_join(alex_nldas, alex_gauge, by = "fips")
a <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Alex, 2010")
a

county_range <- compare_rain %>% 
        group_by(fips) %>%
        summarize(lowest_rain = min(rain_mm),
                  highest_rain = max(rain_mm),
                  nldas_rain = mean(tot_precip))
b <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm)) + 
        geom_segment(data = county_range, 
                     aes(x = nldas_rain, xend = nldas_rain,
                         y = lowest_rain, yend = highest_rain,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_rain$tot_precip, compare_rain$rain_mm)) +
        ylim(range(compare_rain$tot_precip, compare_rain$rain_mm)) + 
        ggtitle("Alex, 2010") + 
        xlab("NLDAS-2 rainfall (mm)") + 
        ylab("MWR rainfall reports (mm)")
b
```

```{r echo = FALSE, fig.width = 5.5, fig.height = 4}
compare_rain %>% group_by(fips) %>%
        summarize(ave_rain_mm = mean(rain_mm),
                  ave_tot_precip = mean(tot_precip),
                  n = n(),
                  within_500_km = first(storm_dist) <= 500) %>%
        ggplot(aes(x = ave_tot_precip, y = ave_rain_mm, 
                   size = n, color = within_500_km)) + 
        geom_hline(yintercept = 75, color = "gray") + 
        geom_vline(xintercept = 75, color = "gray") + 
        geom_point(alpha = 0.5) + 
        theme_bw() +
        xlim(c(10, 225)) + ylim(c(10, 225)) + 
        geom_abline(intercept = 0, slope = 1) + 
        xlab("Rainfall based on NLDAS-2 data (mm)") + 
        ylab("Rainfall reported in Monthly \nWeather Report (mm)")  + 
        ggtitle("Alex, 2010") + 
        scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black"), 
                           guide = "none") 
```

```{r echo = FALSE, fig.height = 4, fig.width = 6}
#allison_gauge <- read.csv("~/Documents/hurricaneexposure/writing/DraftExposurePaper/MWR_gauge_rain_data/allison_mwr.csv")
allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv")
allison_nldas <- county_rain(counties = unique(allison_gauge$fips),
                             start_year = 2001, end_year = 2001,
                             rain_limit = 0, dist_limit = 1500,
                             days_included = c(-2, -1, 0, 1))
allison_nldas <- allison_nldas %>%
        filter(storm_id == "Allison-2001")
allison_gauge$fips <- as.character(allison_gauge$fips)
compare_rain <- right_join(allison_nldas, allison_gauge, by = "fips")

a <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Allison, 2011")
a
county_range <- compare_rain %>% 
        group_by(fips) %>%
        summarize(lowest_rain = min(rain_mm),
                  highest_rain = max(rain_mm),
                  nldas_rain = mean(tot_precip))
b <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm)) + 
        geom_segment(data = county_range, 
                     aes(x = nldas_rain, xend = nldas_rain,
                         y = lowest_rain, yend = highest_rain,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_rain$tot_precip, compare_rain$rain_mm)) +
        ylim(range(compare_rain$tot_precip, compare_rain$rain_mm)) + 
        ggtitle("Allison, 2001") + 
        xlab("NLDAS-2 rainfall (mm)") + 
        ylab("MWR rainfall reports (mm)")
b
```

```{r echo = FALSE, fig.width = 5.5, fig.height = 4}
compare_rain %>% group_by(fips) %>%
        summarize(ave_rain_mm = mean(rain_mm),
                  ave_tot_precip = mean(tot_precip),
                  n = n(),
                  within_500_km = first(storm_dist) <= 500) %>%
        ggplot(aes(x = ave_tot_precip, y = ave_rain_mm, 
                   size = n, color = within_500_km)) + 
        geom_hline(yintercept = 75, color = "gray") + 
        geom_vline(xintercept = 75, color = "gray") + 
        geom_point(alpha = 0.5) + 
        theme_bw() +
        xlim(c(25, 200)) + ylim(c(25, 800)) + 
        geom_abline(intercept = 0, slope = 1) + 
        xlab("Rainfall based on NLDAS-2 data (mm)") + 
        ylab("Rainfall reported in Monthly \nWeather Report (mm)") + 
        ggtitle("Allison, 2001") + 
        scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black"), 
                           guide = "none") 
```

```{r echo = FALSE, fig.height = 4, fig.width = 6}

load("ike_gauge.RData")
ike_gauge$rain_mm <- convert_precip(ike_gauge$rain_in, 
                                    "inches", "mm")
ike_gauge$fips <- sprintf("%05s", ike_gauge$fips)
ike_nldas <- county_rain(counties = unique(ike_gauge$fips),
                             start_year = 2008, end_year = 2008,
                             rain_limit = 0, dist_limit = 1500,
                             days_included = c(-2, -1, 0, 1))
ike_nldas <- ike_nldas %>%
        filter(storm_id == "Ike-2008")
ike_gauge$fips <- as.character(ike_gauge$fips)
compare_rain <- right_join(ike_nldas, ike_gauge, by = "fips")
a <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Ike, 2008")
a

county_range <- compare_rain %>% 
        group_by(fips) %>%
        summarize(lowest_rain = min(rain_mm),
                  highest_rain = max(rain_mm),
                  nldas_rain = mean(tot_precip))
b <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm)) + 
        geom_segment(data = county_range, 
                     aes(x = nldas_rain, xend = nldas_rain,
                         y = lowest_rain, yend = highest_rain,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_rain$tot_precip, compare_rain$rain_mm)) +
        ylim(range(compare_rain$tot_precip, compare_rain$rain_mm)) + 
        ggtitle("Ike, 2008") + 
        xlab("NLDAS-2 rainfall (mm)") + 
        ylab("MWR rainfall reports (mm)")
b
```

[JMF-- Missing values are those from Arkansas (missing the leading 0 in fips code); GBA: I fixed this. This is ultimately something I need to fix in the `hurricaneexposure` package, but I've got this example fixed to run without losing the FIPS with leading zeros.]

```{r echo = FALSE, fig.width = 5.5, fig.height = 4}
compare_rain %>% group_by(fips) %>%
        summarize(ave_rain_mm = mean(rain_mm),
                  ave_tot_precip = mean(tot_precip),
                  n = n(),
                  within_500_km = first(storm_dist) <= 500) %>%
        ggplot(aes(x = ave_tot_precip, y = ave_rain_mm, 
                   size = n, color = within_500_km)) + 
        geom_hline(yintercept = 75, color = "gray") + 
        geom_vline(xintercept = 75, color = "gray") + 
        geom_point(alpha = 0.5) + 
        theme_bw() +
        xlim(c(10, 225)) + ylim(c(10, 225)) + 
        geom_abline(intercept = 0, slope = 1) + 
        xlab("Rainfall based on NLDAS-2 data (mm)") + 
        ylab("Rainfall reported in Monthly \nWeather Report (mm)") + 
        ggtitle("Allison, 2001") + 
        scale_color_manual(values = c("blue", "black"), guide = "none") 
```

Comparing MWR precipitation data to NLDAS-2 data for Frances-2004:

```{r}
load("frances_gauge.RData")
frances_gauge$rain_mm <- convert_precip(frances_gauge$precip_in, "inches", "mm")
frances_gauge$fips <- sprintf("%05s", frances_gauge$fips)
frances_nldas <- county_rain(counties = unique(frances_gauge$fips),
                             start_year = 2004, end_year = 2004,
                             rain_limit = 0, dist_limit = 1500,
                             days_included = c(-2, -1, 0, 1))
frances_nldas <- frances_nldas %>%
        filter(storm_id == "Frances-2004")
frances_gauge$fips <- as.character(frances_gauge$fips)
compare_rain <- right_join(frances_nldas, frances_gauge, by = "fips")
a <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm,
                              color = fips)) +
        geom_point() + 
        theme_bw() + 
        ggtitle("Frances, 2004")
a

county_range <- compare_rain %>% 
        group_by(fips) %>%
        summarize(lowest_rain = min(rain_mm),
                  highest_rain = max(rain_mm),
                  nldas_rain = mean(tot_precip))
b <- ggplot(compare_rain, aes(x = tot_precip, y = rain_mm)) + 
        geom_segment(data = county_range, 
                     aes(x = nldas_rain, xend = nldas_rain,
                         y = lowest_rain, yend = highest_rain,
                         group = fips), color = "gray") + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        geom_abline(aes(intercept = 0, slope = 1)) + 
        xlim(range(compare_rain$tot_precip, compare_rain$rain_mm)) +
        ylim(range(compare_rain$tot_precip, compare_rain$rain_mm)) + 
        ggtitle("Frances, 2004") + 
        xlab("NLDAS-2 rainfall (mm)") + 
        ylab("MWR rainfall reports (mm)")
b
```

```{r echo = FALSE, fig.width = 5.5, fig.height = 4}
compare_rain %>% group_by(fips) %>%
        summarize(ave_rain_mm = mean(rain_mm),
                  ave_tot_precip = mean(tot_precip),
                  n = n(),
                  within_500_km = first(storm_dist) <= 500) %>%
        ggplot(aes(x = ave_tot_precip, y = ave_rain_mm, 
                   size = n, color = within_500_km)) + 
        geom_hline(yintercept = 75, color = "gray") + 
        geom_vline(xintercept = 75, color = "gray") + 
        geom_point(alpha = 0.5) + 
        theme_bw() +
        xlim(c(10, 225)) + ylim(c(10, 225)) + 
        geom_abline(intercept = 0, slope = 1) + 
        xlab("Rainfall based on NLDAS-2 data (mm)") + 
        ylab("Rainfall reported in Monthly \nWeather Report (mm)") + 
        scale_color_manual(values = c("blue", "black"), guide = "none") 
```

####How does NLDAS-2 data compare with rnoaa data from the `countyweather` package in R?

```{r eval = FALSE}
options("noaakey" = Sys.getenv("noaakey"))
#Alex-2010
alex_fips <- alex_nldas$fips
alex_fips <- unique(alex_fips)
```
```{r}
#Ike-2008
ike_fips <- ike_nldas$fips
ike_fips <- unique(ike_fips)
```
```{r}
#Allison-2010
allison_fips <- allison_nldas$fips
allison_fips <- unique(allison_fips)
```
```{r}
#Frances-2004
frances_fips <- frances_nldas$fips
frances_fips <- unique(frances_fips)
```

@@

As of right now, I am not going to do Frances (there are 184 unique fips).

Create a timeseries for the top 15 FIPS by reporting stations for Alex and Ike:

[JMF-- It looks like precipitation measurements from `countyweather` are already in mm]

>>>>

# Pull in rain data and limit to storm-county pairs with 150 mm or more 
# of rainfall over lag -5 to 3. Figure out the cumulative values at each 
# day within a county-storm for each lag day.
rain_lag_1 <- rbind(rain, lag_6) %>%
        arrange(storm_id, fips, lag) %>%
        mutate(storm_county = paste0(storm_id, "_", fips)) %>%
        filter(storm_county %in% storm_lag$storm_county) %>%
        group_by(storm_county) %>%
        mutate(cumulative_rain = cumsum(precip),
               next_lag = lead(lag),
               next_cumulative_rain = lead(cumulative_rain),
               over_50 = lead(precip) >= 50)
rain_lag <- rain_lag_1 %>% filter(lag != 3)
high_rain <- rain_lag %>%
        ungroup() %>%
        group_by(lag) %>%
        summarize(n_high = sum(over_50)) %>%
        ungroup() %>%
        mutate(lag = lag + 1)
to_highlight <- data.frame(x = c(-3.1, -3.1, 1.1, 1.1) + 0.5, 
                           y = c(-15, 500, 500, -15))

@@

```{r daily_rain, echo = FALSE, fig.width = 6, fig.height = 4, fig.align = "center", fig.cap = "Boxplots of daily rainfall for days at a lag of five days before (-5) to three days after (3) compared to the date of the storm's closest approach to a county. These values are only for county-storm observations for which a total of at least 150 millimeters of rain fell over the full period (lags -5 to 3)."}
rain_lag_1 %>%
        filter(lag != -6) %>%
        ggplot(aes(x = lag, y = precip, group = lag)) + 
        geom_boxplot(outlier.size = 0.5) + theme_few() + 
        xlab("Lag period from storm's closest approach to county") + 
        ylab("Daily rainfall in county (mm)") 
```

@@

The method we use to assess storm-related rain exposure is spatially and temporally complete, so it allows us to assess storm exposure for every eastern US county for every storm from 1988 to 2011. However, we wanted to compare this method of assessing storm-related rain with other ways of measuring storm-related rain. We compared our method with two other assessments. First, we compared our results to rain totals from ground-based monitors, using county-level values averaged across all available ground-based precipitation monitors in the county. For this assessment, we picked several counties and compared storm-related rain estimates using our data and ground-based monitor estimates for all storms over our study period. We also picked several storms and compared our method to values from ground-based monitors for several counties for that storm.

Second, researchers from the [NWS? Hurricane Center?] publish yearly tropical cyclone season summaries in the journal *Monthly Weather Review*. For some storms, this report includes multiple ground-based measurements of total precipitation and wind speeds (both sustained and gusts). We picked a few storms with multiple ground-based measurements included in the *Monthly Weather Review* report, and compared assessments using our method to these reported storm-related measurements.

Here are results for comparing our method to using ground-based monitors. First, we looked at several counties, comparing estimates of storm-associated rain using NLDAS-2 versus ground-based monitors (Figure [x]).

[Add in figures by county here for county averages of ground-based monitors versus our method.]

Next, we looked at several different counties within a few notable storms. These storms were picked based on storms that had a large number of ground-based precipitation observations in either the main text or supplemental material of the *Monthly Weather Review* report.

[Move figures here for NLDAS vs. county average values.]

Finally, we compared exposure results based on the NLDAS-2 precipitation data with values from monitors reported for certain storms in the *Monthly Weather Reviews* yearly summary of the Atlantic basin tropical storm season (Figure \ref{fig:mwr_comparison}). Here are some of the plots comparing county-level estimates of rain exposure from the NLDAS-2 data to all *MWR*-reported rainfall estimates within the county:

```{r mwr_comparison, fig.width = 10, fig.height = 10, echo = FALSE, fig.align = "center", fig.cap = "Comparison of estimated storm-associated rainfall from our method (using NLDAS-2 data for a window from two days before to one day after the storm's closest approach to the county) with ground-based observations reported in the Monthly Weather Review report. Observations within the same county are linked by a line segment. Blue points / lines indicate a county for which the storm came within 500 kilometers, while gray points / lines indicate that the storm did not come within 500 kilometers of the county (and so would be excluded as an 'exposed' county based on distance). The diagonal gray line shows where points would fall if assessments were identical. Horizontal and vertical gray lines show the 75 millimeter threshold for storm-associated rainfall we use for rain-related storm exposure classification."}
# Function to plot comparison between two datasets
plot_mwr_compare <- function(data){
        storm_year <- as.numeric(gsub(".+-", "", data$storm_id[1]))
        data$fips <- sprintf("%05s", data$fips)
        storm_title <- paste(unlist(strsplit(as.character(data$storm_id[1]), 
                                       split = "-")), collapse = ", ")

        compare_rain <- county_rain(counties = unique(data$fips),
                             start_year = storm_year, 
                             end_year = storm_year,
                             rain_limit = 0, dist_limit = 2000) %>%
                dplyr::filter(storm_id == data$storm_id[1]) %>%
                dplyr::right_join(data, by = "fips") %>%
                dplyr::mutate(within_500_km = storm_dist <= 500)

        county_range <- compare_rain %>% 
                group_by(fips) %>%
                summarize(lowest_rain = min(rain_mm),
                  highest_rain = max(rain_mm),
                  nldas_rain = mean(tot_precip),
                  within_500_km = first(storm_dist) <= 500)

        plot_mwr <- ggplot(compare_rain, 
                   aes(x = tot_precip, y = rain_mm,
                       color = within_500_km)) + 
        geom_hline(yintercept = 75, color = "lightgray") + 
        geom_vline(xintercept = 75, color = "lightgray") + 
                geom_abline(aes(intercept = 0, slope = 1), color = "gray") + 
        geom_segment(data = county_range, 
                     aes(x = nldas_rain, xend = nldas_rain,
                         y = lowest_rain, yend = highest_rain,
                         group = fips, color = within_500_km)) + 
        geom_point(aes(group = NA), alpha = 0.5) + 
        ggthemes::theme_few() + 
        xlim(range(compare_rain$tot_precip, compare_rain$rain_mm)) +
        ylim(range(compare_rain$tot_precip, compare_rain$rain_mm)) + 
        ggtitle(storm_title) + 
        xlab("NLDAS-2 rainfall (mm)") + 
        ylab("Monthly Weather Report\n rainfall reports (mm)") + 
        scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "black"), 
                           guide = "none") 
        return(plot_mwr)
}

allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv") 
allison_mwr <- plot_mwr_compare(allison_gauge)

load("frances_gauge.RData")
frances_gauge$rain_mm <- convert_precip(frances_gauge$precip_in,
                                        "inches", "mm")
frances_mwr <- plot_mwr_compare(frances_gauge)

load("ike_gauge.RData")
ike_gauge$rain_mm <- convert_precip(ike_gauge$rain_in, 
                                    "inches", "mm")
ike_mwr <- plot_mwr_compare(ike_gauge)

load("alex_gauge.Rdata")
alex_mwr <- plot_mwr_compare(alex_gauge)

grid.arrange(allison_mwr, frances_mwr, ike_mwr, alex_mwr, ncol = 2)
```

```{r eval = FALSE}
options("noaakey" = Sys.getenv("noaakey"))
#Alex-2010
alex_fips <- alex_nldas$fips
alex_fips <- unique(alex_fips)

#Ike-2008
ike_fips <- ike_nldas$fips
ike_fips <- unique(ike_fips)

#Allison-2010
allison_fips <- allison_nldas$fips
allison_fips <- unique(allison_fips)

#Frances-2004
frances_fips <- frances_nldas$fips
frances_fips <- unique(frances_fips)
```

@@
